Program Equivalence : A Relational Separation Logic interactive prover implemented in Maude

### 2018-06-30 paper/paper_after_guide.tex
added new figure for states

\tableofcontents
\pagebreak
\begin{abstract}
	Lucrarea de față descrie procesul de dezvoltare al unui utilitar interactiv utilizat în analiza formală a programelor, folosindu-se de Logica Separatoare Relațională \cite{relational} și Logica Separatoare \cite{primer} \cite{SeparationLogic}, utilizând Maude \cite{maudesite} drept cadru de dezvoltare. Componenta interactivă se îmbină cu o componentă automatizată, prin care părți din demonstrație vor fi realizate automat. Lucrarea este orientata pe rezolvarea problemei prin intermediul funcționalităților oferite de Maude și va include secvențe de cod ce vor ilustra conceptele cheie legate de modelarea si aplicarea fundamentelor teoretice utilizate pe parcursul proiectului.
\end{abstract}
\section*{Introduction}
The present paper describes the development of an interactive tool for reasoning how two programs are related, based theoretical concepts from the field of formal program analysis and technologies which facilitate the implementation of those concepts. \\

The tool represents an implementation of Hoare Logic - which allows formal reasoning about a program - along  with 2 of its extensions, namely the Separation Logic and Relational Separation Logic \cite{relational}. The 2 extensions simplify the Hoare Logic proofs, mainly using the "*" connector, allowing for local reasoning about the effects of statements in a program . The tool has been implemented in Maude, a high performance logical framework with powerful metalanguage applications which facilitate the implementations of executable environments for logics.\\

The tool is built as a CLI which helps with reasoning about how two programs are related using Relational Separation Logic specifications. As a consequence of the dependency of Relational Separation Logic on Separation Logic, proofs about single programs using the latter are also supported by the tool. The tool has been developed with extensibility in mind, the main desired extensions being concurrent programs support and automatic proofs. \\

The rest of the paper is organized as follows: 
\begin{itemize}
	\item {\textbf{Section 1} will describe the main personal contributions to the realization of this project}
	\item {\textbf{Section 2} will present the problem this project aims to solve}
	\item {\textbf{Section 3} will shortly present some other projects related to ours}
	\item {\textbf{Section 4} will briefly touch upon the theoretical foundations of this project, together with the technologies used to implement them}
	\item {\textbf{Section 5}, the main section of this paper, will present our approach to solving the problem, and discuss how it can be improved in the future.}
\end{itemize}
\section{Contributions}
Personal contributions to the realization of the project : 
\begin{itemize}
	\item Modelled the Relational Logic and Separation Logic using Maude equational and rewriting logic specifications . 
	\item Developed an interactive tool for reasoning about program behaviour using the aforementioned logics.
	\item Automation of tasks, which makes the tool more convenient to use .
	\item Example of formal proof done using the tool - Appendix \ref{proofExample}
\end{itemize}
\section{Description of the problem}
The problem this project is aiming to solve is related to formal reasoning about the execution of code, mainly focusing on how two programs are related to each other (most often the relation to be proven is equivalence)\\

Comparing programs or code fragments and studying their equivalence is part of every software engineer's activities when they are testing an alternative implementation for an existing solution, fixing bugs, launching new product versions, etc . Naturally, for every process completed manually there are efforts being made in order to make it more efficient, less error-prone and, in the end, automate the process all together. Once such a task is automated in software engineering, it can be included in the flow of any research or development phase of a product. An example benefiting from a formal proof of program equivalence is compiler optimization, where the optimized code needs to be equivalent to the input one \cite{relational} . \\

This project aims to lay the foundations of a tool which facilitates formal reasoning about the way two programs are related with a focus on extensibility and automation of tasks.

@@

Several other tools are presented or referenced in \cite{primer} .

\section{Theoretical Foundations and technologies}
This section will briefly introduce the concepts on which the project is based, together with the technologies used to implement it, along with a few references to resources which explain them in depth.
\subsection{Separation Logic}
Separation Logic allows for reasoning about the effects of code on the program state in a formal manner. The main abstraction used in separation logic is the \textbf{Hoare Triple}
\[ \{P\}\  C\  \{Q\} \]
where \(P\) denotes the a condition satisfied by state of the prover before the execution of the command \(C\) and \(Q\) denotes another condition satisfied by state of the prover after the execution of \(C\).
\\

Good references for reading on separation logic include : 
\begin{itemize}
	\item \cite{primer} - for a good introduction and some interesting usages
	\item \cite{SeparationLogic} - for a more in depth explanation, including the mathematical fundaments of the logic
\end{itemize} 
\subsection{Relational Separation Logic}
Relational Separation Logic, which is the central concept of our project builds upon the Separation Logic to reason about how to programs are related, by using the concept of a \textbf{Hoare Quadruple}
\[\{R\}\ {C_1 \atop C_2}\   \{T\}\]
\(R\) represents a \textbf{relation} between the two program states holding before the execution of commands \(C_1\) and \(C_2\) respectively while \(T\) represents a \textbf{relation} which holds after the execution of the two commands. 
\\

@@

 Good references for Maude language include:
 \begin{itemize}
 	\item Maude Primer \cite{primer} which was written for Maude 2.0.1 but still applies to the latest versions of Maude is a good starting point for learning Maude, as it introduces the language in a more informal, friendlier way
 	\item Maude Manual \cite{manual} which presents the whole Maude system in depth, along with the mathematical foundations
 	\item All about Maude book \cite{allAboutMaude} which includes everything contained in the manual and presents some of the more relevant tools implemented in Maude .
 \end{itemize}
\section{Relational Separation Logic Interactive Prover} 
This section will include details related to the implementation and usage of the Relational Separation Logic prover and highlight a few of Maude's features which made it a language fit for the purpose of this project.
\\

The information presented in this section assumes the reader's familiarity with Maude, especially the META-LEVEL module and with the underlying theoretical concepts implemented by the prover (Separation Logic and Relational Separation Logic). References for these can be found in the previous section.
\\

This section will include snippets of Maude code, but since Maude is a very expressive and self-explanatory language, I have not included of thorough presentation of every language construct defined for the scope of this project. Ideas presented will be accompanied by some examples and key parts of the implementation but for a more detailed presentation, please refer to the code itself.
\subsection{Language}
The prover supports expressing programs in a simple, imperative language, commonly used throughout papers  \cite{primer} \cite{SeparationLogic} related to the subject of program verification, including the one describing Relational Separation Logic \cite{relational}, around  which this project is based. 
\subsubsection{Storage Model}
A state of in our storage model is defined by a pair consisting of a \(Store\) and a \(Heap\) .
\\
Assuming that all the variables usable in programs from the set \(\mathtt{Vars}\) and the set of positive natural numbers is denoted by \(\mathtt{PosNats}\): \\

A \(Store\) is defined as:
\[ \mathtt{S} : \mathtt{Vars} \rightarrow \mathtt{PosNats} \]

A \(Heap\) represents a mapping from the \(\mathtt{PosNats}\) to \(\mathtt{Integers}\)
\[ \mathtt{H} : \mathtt{PosNats} \rightarrow \mathtt{Integers} \] 
More informally, the \(Store\) holds the value of the variables while the \(Heap\) maps the active memory cells during a program execution to their contents.

@@

		Integer Expressions& \(E ::= x \mid Integer \mid E\ \mathtt{plus}\ E \mid E\ \mathtt{times}\ E \mid E\ \mathtt{minus}\ E \) \\

@@

		Boolean Expressions& \(B ::= false \mid true \mid B\ \mathtt{\&\&}\ B \mid  B\ ||\ B \mid B\  \mathtt{->}\ B \mid B \mathtt{<=>} B \mid \ ! B \mid E\ \mathtt{ge}\ E \mid E\ \mathtt{le}\ E \mid E\ \mathtt{eqs}\ E \)\\

@@

The syntax of the language is presented in Figure \ref{fig:langSyntax} . It represents an adapted subset of the language presented in the Relational Separation Logic paper \cite{relational} . 
\paragraph{Semantics of the language}

@@

	The semantics of the previously defined language are standard, but a few clarifications are necessary to point out how our language constructs interact with the storage model, as well as a few differences from regular programming languages:

@@

		\item{Verbosity of operators - most operators are replaced by their literal names ( \(+\) becomes \(\mathtt{plus}\) etc.); we have opted for this approach to avoid complication when implementing the language in Maude, because most of the operators are already defined in Maude for its built-in types, and overloading them could cause conflicts since we are basing our own defined types on Maude's primitives.}
		\subitem{\( B\ \mathtt{ge}\ B\) translates to the usual \(>=\) operator}
		\subitem{\( B\ \mathtt{le}\ B\) translates to the usual \(<=\) operator}
		\subitem{\( B\ \mathtt{eqs}\ B\) translates to the usual \(==\) equality comparison operator}
		\subitem{\( B\ \mathtt{->}\ B\) denotes implication between boolean expressions}
		\subitem{\(B\ \mathtt{<=>}\ B\) denotes equivalence between boolean expressions}
		\subitem{The rest of the operators are mostly similar to their C++ or Java counterparts and their semantics are self explanatory}

@@

		\subitem{\(x := alloc(E)\) allocates a new cell in the memory, initializes it with the value of \(E\) and stores its address in the variable \(x\)}
		\subitem{\(x := free(E)\) deallocates the cell at the address equal to the value of \(E\)}

@@

		\subitem{\( x := [E] \) reads the contents of the memory cell at address \(E\) and stores the value in the variable \(x\)}
		\subitem{\( [E] := E\prime \) updates the contents of the memory cell at address \(E\) with the value of the expression \(E\prime\)}
		\subitem{\( x := E \) updates the value of the variable \(x\) with the value of the expression \(E\) - note that this command does not modify the heap in any way, as it usually happens with regular programming languages when updating the value of a variable}

@@

	\paragraph{Maude Modelling of the language}
	Maude makes the representation of programming languages grammar a straight forward and natural process, by defining sorts corresponding to language elements and their associated operators. The sorts defined in Maude for the grammar are the following : 

@@

		\item \texttt{AExp} - abbreviation for Arithmetic Expression - equivalent for Integer expressions in the specification. To capture the first two possible forms of Integer expressions in Figure \ref{fig:langSyntax}, the AExp sort is a super sort of \texttt{Int} of \texttt{INT} standard module and \texttt{Id}, which is the sort defined for variables.
		\item \texttt{BExp} - abbreviation for Boolean Expressions - equivalent for the Boolean Expressions type in the specification. It is a super sort of \texttt{Bool} of standard module \texttt{BOOL}. 
		\item \texttt{Command} sort maintains the same name as in the specification.

@@

	An example of operator translation from specification to Maude code is presented in Listing \ref{code:languageGrammer}. Note that we have opted to skip the attributes associated with the operators, as they are irrelevant for the purpose of this example.

@@

	\paragraph{Executable semantics of the language, in Maude}
	Maude also supports specifying executable semantics for the language using rewriting logic \cite{rewrConcurrency}. A proof of concept for a language similar to the one used by the prover is present in the code of the project (\textbf{simpleLanguage.maude}), but it is out of scope for the presentation of the prover . An executable environment for the programming language used in the prover is a valid candidate for future improvements.

\subsection{Modelling the Separation Logics}
Thanks to Maude's ability to naturally represent logics, the modelling of the Separation Logic and Relational Separation Logic is mostly reduced to translating the logical formulae into appropriate syntax for Maude, making the representational distance (the amount of information lost when representing a theoretical concept into a programming language)\cite{manual} almost non existent. \\

Each proof rule in the Separation Logic and Relational Separation Logic has been translated into a Maude rewrite rule. 
\\
The prover is centered around the concept of \textbf{Goals} which can be one of the three things: 
\begin{itemize}
	\item {Hoare Quadruple} - corresponding to Relational Separation Logic
	\item {Hoare Triple} - corresponding to Separation Logic
	\item {Implication between relations / assertions} 
\end{itemize} 
The proof rules are interpreted in a bottom-up manner, meaning that, applying a rule to a goal, which must match the conclusion of the used rule, will generate new goals consisting of the hypothesis: Figure \ref{fig:SLTranslation} presents how a Separation Logic rule has been translated into Maude syntax and Figure \ref{fig:RSLTranslation} presents the same process for the Relational Separation Logic counterpart of the rule.
\begin{figure}[h]
	Consequence Rule
	\medskip
	\[
	\begin{prooftree}
	\Hypo{P \Rightarrow P\prime}
	\Hypo{\{P\prime\} C \{Q\prime\}}
	\Hypo{Q\prime \Rightarrow Q}
	\Infer 3 {\{P\}\  C  \{Q\}}
	\end{prooftree}	
	\]
	Is represented in Maude syntax as : 
\begin{lstlisting}[caption=Separation Logic Consequence rule]
	rl [SeparationLogic-Consequence] : {P} C {Q} => ((P => P1) <> ({P1} C {Q1})) <> (Q1 => Q) [nonexec] .
\end{lstlisting}
	\caption{Example of a translation to Maude syntax of a Separation Logic proof rule}
	\label{fig:SLTranslation}
\end{figure}

\begin{figure}[h]
	Consequence Rule
	\[
	\begin{prooftree}
	\Hypo{R \Rightarrow R_1}
	\Hypo{\hq{R_1}{C}{C\prime}{S_1}}
	\Hypo{S_1 \Rightarrow S}
	\Infer 3 {\hq{R}{C}{C\prime}{S}}
	\end{prooftree}	
	\]
	Is represented in Maude syntax as : 
	\begin{lstlisting}[caption=Relational Separation Logic Consequence rule]
	rl [Consequence] : { R } C1 -- C2  { S } => ((R => R1) <> ({R1} C1 -- C2 {S1})) <> (S1 => S) [nonexec] .
	\end{lstlisting}
	\caption{Example of a translation to Maude syntax of a Relational Separation Logic proof rule}
	\label{fig:RSLTranslation}
\end{figure}
The \texttt{<>} operator is used to separate sub-goals generated by the rule.
\\

@@

All the rules are marked with the attribute \texttt{[nonexec]} because we want them to be applied in a controlled manner, at the meta-level. Some of the rules, the examples included, are not even valid executable Maude rules, because they make use of variables before binding them (for example \texttt{P1} is used in a rule before it is bound). At the meta-level however we can manually bind values to those variables, as we will exemplify when discussing how we apply these rules to goals.
\subsubsection{Sort Hierarchy}
Much in the same manner as we modelled the language grammar, the Relational Separation Logic \cite{relational} and Separation Logic \cite{SeparationLogic} \cite{primer} have been modeled by defining sorts and declaring operators using these sorts. The main point of reference for our modelling has been \cite{relational}, for both logics. Some examples of translations from specification to code are included below.
\begin{itemize}
	\item \texttt{Assertion} is the sort used to handle Separation Logic assertions. It is a super sort of the \texttt{BExp} sort defined as part of our language.
	\item \texttt{HoareTriple} is the sort associated with the concept with the same name, in a Separation Logic context.
	\begin{lstlisting}[caption=Hoare Triple constructor operator]
op {_}_{_} : Assertion Command Assertion -> HoareTriple .\end{lstlisting}
	\item \texttt{Relation} is the sort associated with Relational Separation Logic relations. As it was the case with \texttt{Assertion}, it is a super sort of the \texttt{BExp} sort of our language grammar. Because both the \texttt{Relation} and \texttt{Assertion} are super sorts of the same sort, Maude will display a few warnings which are not affecting functionality, but in the future, creating a separate \texttt{BExp} sort for both the \texttt{Assertion} and \texttt{Relation} sorts would be a better option.
	\item \texttt{HoareQuadruple} is again associated with the concept with the same name, in the Relational Separation Logic context.
	\begin{figure}[h]
	\begin{lstlisting}[caption=Examples of Relational Separation Logic specific constructs]
--- Operator which creates a relation from two assertions
op [_;_] : Assertion Assertion -> Relation .
--- Operator which constructs a Hoare Quadruple 
op {_}_--_{_} : Relation Command Command Relation -> HoareQuadruple .\end{lstlisting}
	\end{figure}
\end{itemize}
The other operators for the Assertion and Relation languages have been defined in the same manner as the programming language operators, following their presentation in \cite{relational}.

\subsection{Interaction by Maude LOOP-MODE}
Maude's LOOP-MODE module is the entry point for any interactive Maude application, being the only way to store a state between commands. In this chapter we will discuss how we handle input and output , the structure of a state and the role every element plays in the execution of the prover.\\

We will make a distinction between \textbf{Loop state}, which specific to Maude's LOOP-MODE module and the \textbf{Prover state} which is specific to our use-case, and holds the necessary data related to the state of the proof we are currently working on. The \textbf{Prover state} is encapsulated within \textbf{Loop state}.
\begin{figure}[h]
	\noindent\rule{\linewidth}{0.4pt}

>>>>

\tableofcontents
\pagebreak
\begin{abstract}
	Lucrarea de față descrie procesul de dezvoltare al unui utilitar interactiv utilizat în analiza formală a programelor, folosindu-se de Logica Separatoare Relațională \cite{relational} și Logica Separatoare \cite{primer} \cite{SeparationLogic}, utilizând Maude \cite{maudesite} drept cadru de dezvoltare. Componenta interactivă se îmbină cu o componentă automatizată, prin care părți din demonstrație vor fi realizate automat. Lucrarea este orientata pe rezolvarea problemei prin intermediul funcționalităților oferite de Maude și va include secvențe de cod ce vor ilustra conceptele cheie legate de modelarea si aplicarea fundamentelor teoretice utilizate pe parcursul proiectului.
\end{abstract}
\section*{Introduction}
In the present paper I describe the development of an interactive tool for reasoning about how two programs are related, in a formal manner. \\

The tool is an implementation of Hoare Logic, which allows formal reasoning about a program along with two of its extensions, namely the Separation Logic and Relational Separation Logic \cite{relational}. The two extensions simplify the Hoare Logic proofs, mainly using the "*" connector, allowing for local reasoning about the effects of statements in a program. I have the tool implemented in Maude, a high performance logical framework with powerful metalanguage applications that facilitate the implementation of executable environments for logics.\\

The tool is built as a CLI that helps with reasoning about how two programs are related using Relational Separation Logic specifications. As a consequence of the dependency of Relational Separation Logic on Separation Logic, proofs about single programs using the latter are also supported by the tool. The tool has been developed with extensibility in mind, the main desired extensions being concurrent programs support and automatic proofs. \\

The paper is organized as follows: 
\begin{itemize}
	\item {\textbf{Section 1} describes the main personal contributions to the realization of this project;}
	\item {\textbf{Section 2} presents the problem this project aims to solve;}
	\item {\textbf{Section 3} shortly presents some other projects related to this one;}
	\item {\textbf{Section 4} briefly touches upon the theoretical foundations of this project, together with the technologies used to implement them;}
	\item {\textbf{Section 5}, the main section of this paper, presents my approach to solving the problem, and discuss how it can be improved in the future.}
\end{itemize}
\section{Contributions}
Personal contributions:
\begin{itemize}
	\item Modelling the Relational Separation Logic and Separation Logic using Maude equational and rewriting logic specifications; 
	\item Developing an interactive tool for reasoning about program behaviour using the aforementioned logics;
	\item Automation of tasks that make the tool more convenient to use;
	\item Example of formal proof performed using the tool - Appendix \ref{proofExample}
\end{itemize}
Throughout the development of the prover, I have learned how to use a different, unfamiliar technology, that is a good fit for  modelling theoretical concepts which were also new to me. This learning by applying process properly introduced me to both the technology and the logics involved in formally verifying program equivalence. 
\section{Description of the Problem}
Comparing programs or code fragments and studying their equivalence is part of the activities of every software engineer, when they are testing an alternative implementation for an existing solution, fixing bugs, launching new product versions, etc. Naturally, for every process completed manually there are efforts being made in order to make it more efficient, less error-prone and, in the end, automate the process all together. These goals can be achieved by creating an easy to use and reliable formal  method of verifying program equivalence.
\\

An example benefiting from a formal proof of program equivalence is compiler optimization, where the optimized code needs to be equivalent to the input one \cite{relational}.\\

The problem this project aims to solve is reducing the gap between theoretical and practical aspects of formal proofs for program equivalence by creating an executable environment for Relational Separation Logic in which the user can prove the equivalence of programs. Through the creation of such a tool, the theoretical concepts become more accessible, easier to apply and understand. Making formal verification methods more accessible leads to an overall increase in the quality of software, by replacing the existing methods of verifying equivalence, that are often prone to errors, with more robust ones.\\

The difficult aspects of creating this tool include:
\begin{itemize}
	\item Losing as little generality as possible when modelling the theoretical concepts in a programming language;
	\item Approaching computationally-hard problems required for Relational Separation Logic proofs;
	\item Understanding the complexities of Relational Separation Logic and create a user experience which facilitate the application of the logic.
\end{itemize}

@@

Several other tools are presented or referenced in \cite{primer}.

\section{Theoretical Foundations and technologies}
This section will briefly introduce the concepts on which the project is based, together with the technologies used to implement it, along with a few references to resources which explain them in depth.
\subsection{Separation Logic}
Separation Logic allows for reasoning about the effects of code on the program state in a formal manner. The main abstraction used in separation logic is the \textbf{Hoare Triple}:
\[ \{P\}\  C\  \{Q\} ,\]
where \(P\) denotes the a condition satisfied by state of the prover before the execution of the command \(C\) and \(Q\) denotes another condition satisfied by state of the prover after the execution of \(C\).
\\

Good references for reading on separation logic include: 
\begin{itemize}
	\item \cite{primer} - for a good introduction and some interesting usages ;
	\item \cite{SeparationLogic} - for a more in depth explanation, including the mathematical fundaments of the logic.
\end{itemize} 
\subsection{Relational Separation Logic}
Relational Separation Logic, which is the central concept of our project builds upon Separation Logic to reason about how two programs are related, by using the concept of a \textbf{Hoare Quadruple}:
\[\{R\}\ {C_1 \atop C_2}\   \{T\},\]
where \(R\) represents a \textbf{relation} between the two program states holding before the execution of commands \(C_1\) and \(C_2\), while \(T\) represents a \textbf{relation} that holds after the execution of the two commands. 
\\

@@

 Good references for Maude language include:
 \begin{itemize}
 	\item The Maude Primer \cite{primer}, which was written for Maude 2.0.1 but still applies to the latest versions of Maude is a good starting point for learning Maude, as it introduces the language in a more informal, friendlier way;
 	\item The Maude Manual \cite{manual}, which presents the whole Maude system in depth, along with the mathematical foundations;
 	\item "All about Maude" book \cite{allAboutMaude}, which includes everything contained in the manual and presents some of the more relevant tools implemented in Maude.
 \end{itemize}
\section{Relational Separation Logic Interactive Prover} 
This section includes details related to the implementation and usage of the Relational Separation Logic prover and highlights a few of the features provided by Maude that made it a language fit for the purpose of this project.
\\

The information presented in this section assumes a degree familiarity with Maude, especially the META-LEVEL module, and with the underlying theoretical concepts implemented by the prover (Separation Logic and Relational Separation Logic). References for these can be found in the previous section.
\\

This section includes snippets of Maude code, but since Maude is a very expressive and self-explanatory language, I have not included a complete presentation of every language construct defined for the scope of this project. The ideas presented are accompanied by some examples and key features of the implementation. For a more detailed presentation, please refer to the code itself.
\subsection{Language}
The prover supports expressing programs in a simple, imperative language, commonly used throughout papers  \cite{primer} \cite{SeparationLogic} related to the subject of program verification, including the one describing Relational Separation Logic \cite{relational}, around  which this project is based. 
\subsubsection{Storage Model}
A state of in our storage model is defined by a pair consisting of a \texttt{Store} and a \texttt{Heap}.
\\
The set of program variables is \(\mathtt{Vars}\). The set of positive natural numbers is denoted by \(\mathtt{PosNats}\): 
\begin{flushleft}
A \texttt{Store} is defined as:
\[ \mathtt{S} : \mathtt{Vars} \rightarrow \mathtt{PosNats} .\]

A \texttt{Heap} represents a mapping from the \(\mathtt{PosNats}\) to \(\mathtt{Integers}\) and is defined as:
\[ \mathtt{H} : \mathtt{PosNats} \rightarrow \mathtt{Integers} .\] 
\end{flushleft}
Informally, the \texttt{Store} holds the value of the variables while the \texttt{Heap} maps the active memory cells during a program execution to their contents.

@@

		Integer Expressions& \(E ::= x \mid \textit{Integer} \mid E\ \mathtt{plus}\ E \mid E\ \mathtt{times}\ E \mid E\ \mathtt{minus}\ E \) \\

@@

		Boolean Expressions& \(B ::= \textit{false} \mid \textit{true} \mid B\ \mathtt{\&\&}\ B \mid  B\ ||\ B \mid B\  \texttt{->}\ B \mid B \texttt{<=>} B \mid \ ! B \mid E\ \mathtt{ge}\ E \mid E\ \mathtt{le}\ E \mid E\ \mathtt{eqs}\ E \)\\

@@

The syntax of the language is presented in Figure \ref{fig:langSyntax}. It represents a subset of the language presented in the Relational Separation Logic paper \cite{relational}. 
\paragraph{}

@@

	The semantics of the previously are standard, but a few clarifications are necessary to point out how our language constructs interact with the storage model, as well as a few differences from regular programming languages:

@@

		\item{Verbosity of operators - most operators are replaced by their literal names (\(+\) becomes \(\mathtt{plus}\) etc.). I have opted for this approach to avoid syntactic complication when implementing the language in Maude, because most of the operators are already defined in Maude for its built-in types, and overloading them could cause conflicts since we are basing our own defined types on Maude's primitives.}
		\subitem{\( B\ \mathtt{ge}\ B\) translates to the usual \(\geq\) operator}
		\subitem{\( B\ \mathtt{le}\ B\) translates to the usual \(\leq\) operator}
		\subitem{\( B\ \mathtt{eqs}\ B\) translates to the usual \(=\) equality comparison operator}
		\subitem{\( B\ \texttt{->}\ B\) denotes implication between boolean expressions}
		\subitem{\(B\ \texttt{<=>}\ B\) denotes equivalence between boolean expressions}
		\subitem{The rest of the operators are self-explanatory}

@@

		\subitem{\(x := alloc(E)\) allocates a new memory cells, initializes it with the value of \(E\) and stores its address in the variable \(x\);}
		\subitem{\(free(E)\) deallocates the cell at the address equal to the value of \(E\).}

@@

		\subitem{\( x := [E] \) reads the contents of the memory cell at address \(E\) and stores the value in the variable \(x\);}
		\subitem{\( [E] := E\prime \) updates the contents of the memory cell at address \(E\) with the value of the expression \(E\prime\);}
		\subitem{\( x := E \) updates the value of the variable \(x\) with the value of the expression \(E\). Note that this command does not modify the heap in any way, as it usually happens with regular programming languages when updating the value of a variable.}

@@

	\paragraph{Maude Modelling of the Language}
	We define sorts corresponding to language elements and their associated operators:

@@

		\item \texttt{AExp} for Integer expressions in the specification. To capture the first two possible forms of Integer expressions in Figure \ref{fig:langSyntax}, the AExp sort is a super sort of \texttt{Int} in the standard \texttt{INT} module and \texttt{Id}, which is the sort defined for variables;
		\item \texttt{BExp} for the Boolean Expressions type in the specification. It is a super sort of \texttt{Bool} of standard module \texttt{BOOL}; 
		\item \texttt{Command} for programs.

@@

	An example of a translation from specification to Maude code is presented in Listing \ref{code:languageGrammer}. Note that we have opted to skip the attributes associated with the operators, as they are irrelevant for the purpose of this example.

@@

	\paragraph{Executable semantics of the language, in Maude}
	Maude also supports specifying executable semantics for the language using rewriting logic \cite{rewrConcurrency}. A proof of concept for a language similar to the one used by the prover is present in the code of the project (\textbf{simpleLanguage.maude}), but it is out of scope for the presentation of the prover. An executable environment for the programming language used in the prover is a valid candidate for future improvements.

\subsection{Modelling the Separation Logics}
Thanks to the ability of Maude to naturally represent logics, the modelling of Separation Logic and Relational Separation Logic is mostly reduced to translating the logical formulae into appropriate syntax for Maude, making the representational distance (the amount of information lost when representing a theoretical concept into a programming language) \cite{manual} almost non existent. \\

Each proof rule in the Separation Logic and Relational Separation Logic has been translated into a Maude rewrite rule. The prover is centered around the concept of \textbf{Goals} which can be one of the three things: 
\begin{itemize}
	\item {Hoare Quadruple} - corresponding to Relational Separation Logic;
	\item {Hoare Triple} - corresponding to Separation Logic;
	\item {Implication between relations and assertions}.
\end{itemize} 
The proof rules are interpreted in a bottom-up manner, meaning that, applying a rule to a goal, which must match the conclusion of the used rule, will generate new goals consisting of the hypothesis: Listing \ref{slConseq} presents how a Separation Logic rule has been translated into Maude syntax and Listing \ref{rslConseq} presents the same process for the Relational Separation Logic counterpart of the rule.
\begin{flushleft}
	The Separation Logic rule of \textsf{Consequence}
\end{flushleft}
	\medskip
	\[
	\begin{prooftree}
	\Hypo{P \Rightarrow P\prime}
	\Hypo{\{P\prime\} C \{Q\prime\}}
	\Hypo{Q\prime \Rightarrow Q}
	\Infer 3 {\{P\}\  C  \{Q\}}
	\end{prooftree}	
	\]
	Is represented in Maude syntax as: 
\begin{lstlisting}[caption=Separation Logic Consequence rule,label=slConseq]
	rl [SeparationLogic-Consequence] : {P} C {Q} => ((P => P1) <> ({P1} C {Q1})) <> (Q1 => Q) [nonexec] .
\end{lstlisting}
\begin{flushleft}
	The Relational Separation Logic rule of \textsf{Consequence}
\end{flushleft}
	\[
	\begin{prooftree}
	\Hypo{R \Rightarrow R_1}
	\Hypo{\hq{R_1}{C}{C\prime}{S_1}}
	\Hypo{S_1 \Rightarrow S}
	\Infer 3 {\hq{R}{C}{C\prime}{S}}
	\end{prooftree}	
	\]
	is represented in Maude syntax as: 
	\begin{lstlisting}[caption=Relational Separation Logic Consequence rule,label=rslConseq]
	rl [Consequence] : { R } C1 -- C2  { S } => ((R => R1) <> ({R1} C1 -- C2 {S1})) <> (S1 => S) [nonexec] .
	\end{lstlisting}
The \texttt{<>} operator is used to separate sub-goals generated by the rule.
\\

@@

All the rules are given the attribute \texttt{[nonexec]} because we want them to be applied in a controlled manner, at the meta-level. Some of the rules, including the examples, are not even valid executable Maude rules, because they make use of variables before binding them For example, in the Separation Logic rule of \textsf{Consequence}, \texttt{P1} is used in a rule before it is bound. At the meta-level, however we can manually bind values to those variables, as we will show when discussing how we apply these rules to goals.
\subsubsection{Sort Hierarchy}
Much in the same manner as we modelled the language grammar, Relational Separation Logic and Separation Logic have been modelled by defining sorts and declaring operators using these sorts. The main point of reference for our modelling has been \cite{relational}, for both logics. Some examples of translations from specification to code are included below.
\begin{itemize}
	\item \texttt{Assertion} is the sort used to handle Separation Logic assertions. It is a super sort of the \texttt{BExp} sort defined as part of our language;
	\item \texttt{HoareTriple} is the sort for Separation Logic \textbf{Hoare Triples} (Listing \ref{HT}):
	\begin{lstlisting}[caption=Hoare Triple constructor operator,label=HT]
op {_}_{_} : Assertion Command Assertion -> HoareTriple .\end{lstlisting}
	\item \texttt{Relation} is the sort for Relational Separation Logic relations. As with \texttt{Assertion}, it is a super sort of the \texttt{BExp} sort of our language. Because both the \texttt{Relation} and \texttt{Assertion} are super sorts of the same sort, Maude will display a few warnings which are not affecting functionality, but in the future, creating a separate \texttt{BExp} sort for both the \texttt{Assertion} and \texttt{Relation} sorts would be a better option.
	\item \texttt{HoareQuadruple} denotes the concept with the same name, in the context of Relational Separation Logic .\\
	Listing \ref{HQ} language constructs relevant for Relational Separation Logic.
	\begin{figure}[h]
	\begin{lstlisting}[caption=Examples of Relational Separation Logic specific constructs,label=HQ]
--- Operator which creates a relation from two assertions
op [_;_] : Assertion Assertion -> Relation .
--- Operator which constructs a Hoare Quadruple 
op {_}_--_{_} : Relation Command Command Relation -> HoareQuadruple .\end{lstlisting}
	\end{figure}
\end{itemize}
The other operators for the Assertion and Relation languages have been defined in the same manner as the programming language operators, following their presentation in \cite{relational}.

\subsection{Interaction by Maude LOOP-MODE}
Our Relational Separation Logic prover is implemented as an interactive CLI program, based on the LOOP-MODE module of Maude, which is the entry point for any interactive Maude application. The LOOP-MODE module provides the only way to store a state between commands. In this chapter we will discuss how we handle input and output, the structure of a state and the role every element plays in the execution of the prover.\\

We will make a distinction between \textbf{Loop state}, which is specific to Maude and the \textbf{Prover state}, which is specific to our use-case, and holds the necessary data related to the state of the proof we are currently working on. The \textbf{Prover state} is encapsulated within \textbf{Loop state} (see Figure \ref{fig:states}).
\begin{figure}[h]
	\includegraphics[width=\textwidth]{"pictures/States".png}
	\caption{Loop and Prover State structure}
	\label{fig:states}
\end{figure}
\subsection{State of the Prover}
Following the examples in the Maude Manual \cite{manual}, our prover state (Figure \ref{fig:states}) is composed of multiple sections modified accordingly during the execution of the prover:
\begin{itemize}
	\item{\texttt{Action} - represents the current action to be executed by the prover;}
	\item{\texttt{Root Goal} - represents the end-goal of the proof;}
	\item{\texttt{Current Goal} - the goal being treated currently - all actions will be applied to this goal;}
	\item{\texttt{Goal Stack} - stack of goals necessary to be proven in order to prove the root goal. The current goal is the last element popped from the stack and new elements are generated and pushed by the application of proof rules to the current goal;}
	\item{\texttt{Proven Goals} - list of goals that have already been proven, or axioms. Each goal will be first checked to see if it matches any goal in this list to prevent repeating the same proofs;}
	\item{\texttt{Staging Output} - contains the output generated by the current action, before being passed to the Loop state. This staging mechanism enables each rewriting rule to generate output independent of the others.}
\end{itemize}
\subsection{Input and Output Handling}
To understand how our prover handles input and output, it is important to note that the flow of the LOOP-MODE module is as follows: 
\begin{enumerate}
	\item User input is placed in the \(\mathtt{INPUT}\) member of the loop state;
	\item Maude engine applies all the possible rewritings, resulting in a new loop state. All the logic of the prover is implemented here, in the form of rewriting rules which modify the prover state;
	\item The member \(\mathtt{OUTPUT}\) of the new loop state is taken and displayed at the console.
\end{enumerate}
There are two main rules used for handling input and output: 
\begin{enumerate}
	\item{The \texttt{in} rule (Listing \ref{inRule}), which is responsible for parsing user input and modifying the action in the prover state accordingly;}
	\begin{lstlisting}[caption=The \texttt{in} rule,label=inRule]
crl [in] : 
--- [ INPUT, <PROVER_STATE>, OUTPUT]
[QIL, < noRule ; RG ; G ; GS ; GL ; nil > , QIL']
=> 
if T:ResultPair? :: ResultPair  
--- If parsing succeeded, place the parsed action in
--- the prover state
then [nil, < downTerm(getTerm(T:ResultPair?), noAction) ; RG ; G ; GS ; GL ; nil >, QIL']  
--- If the parsing failed, display an error
else [nil, < noRule ; RG ; G ; GS ; GL ; nil >, 'ERROR QIL]  
fi  
--- Try to parse the input as an term of sort Action
if QIL =/= nil /\ T:ResultPair? := metaParse(upModule('PROVER-GRAMMAR, false), QIL, 'Action) .
\end{lstlisting}
	\item{The \texttt{out} rule (Listing \ref{outRule}), which takes the staging output from the prover state and appends it to the currently existing output in the loop state.}
	\begin{lstlisting}[caption=The \texttt{out} rule,label=outRule]
crl [out] :
[QIL, < A ; RG ; G ; GS ; GL ;  QIL' >, QIL'']
=> [QIL, < A ; RG ; G ; GS ; GL ; nil >, QIL''  QIL']
if QIL' =/= nil .
\end{lstlisting}
\end{enumerate}
